---
title: "analysis for iLAM paper"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r, quiet=TRUE}
library(behavr)
library(damr)
library(scopr)
library(sleepr)
library(ggetho)
library(zeitgebr)
library(tidyr)
library(zoo)
library(ggpubr)
library(ggplot2)
library(signal)
library(pracma)
library(scales)
library(reshape2)
library(svglite)
library(readr)

if(any(grepl("package:plyr", search()))) detach("package:plyr") else message("plyr not loaded")
library('dplyr')

setwd(dirname(rstudioapi::getSourceEditorContext()$path))

#read in iLAM functions
for (f in list.files("../scripts", pattern="*.R",
                     full.names = TRUE)) {
  source(f)
}

```

Read in the iLAM output for Photinus cages. Plot activity.
```{r}
metadata_p = read_delim("metadata_photinus_v4.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

#link entire metadata with the iLAM files
metadata_entire = metadata_p %>% link_dam_metadata(result_dir = getwd()) 

iLAM_photinus <- load_dam(metadata_entire) 
summary(iLAM_photinus)

#Note that the time variable is in seconds, so t/60 converts to minutes
head(iLAM_photinus$t/60)

#add uid for specific identification for each cage/experiment/species
iLAM_photinus[, uid := 1:.N, meta=T]

#For each timepoint, collapse rows/observations into 30 minute bins
#The sum of every 15 rows 15 = seq(0,14) = 30min bins
iLAM_photinus[, activity_bin := rollapply(activity, list(seq(0, 14)),
                                          FUN=function(x) sum(x, na.rm=TRUE),
                                          partial = TRUE, fill = NA),
              by = xmv(uid)]

#For each uid, retains only the 15th row (t) = 30min bins
iLAM_photinus = iLAM_photinus[, .SD[seq(1, .N, 15)], by = id]

#Calculate average activity across all bins to normalize activity_bins between cages
iLAM_photinus[, uid := 1:.N, meta=T]
iLAM_photinus[, fctr := .(1/mean(activity_bin)), by = xmv(uid)]
iLAM_photinus[, .N, by=xmv(uid)] #verify that proper number of bins were kept

#Multiple activity_bin by normalization factor
iLAM_photinus[, acn := activity_bin*fctr]

#check, how many missing values are there?
table(is.na(iLAM_photinus$activity_bin))

#remove first day of acclimation
iLAM_photinus = iLAM_photinus[t>=days(1)]
iLAM_photinus$t = iLAM_photinus$t-days(1)

#visualize average across individuals within a species for first 72 hours
p_photinus_activity = ggetho(iLAM_photinus,
                             aes(y=acn, colour=species),
                             time_wrap = hours(24.1)) +
  stat_pop_etho(size=0.75) +
  stat_ld_annotations(l_duration = as.numeric(hours(16)),
                      ld_colours = c("#FDFD66","black"))  +
  theme_bw() + theme(text = element_text(size = 12, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.14,0.70),
                     legend.text = element_text(face = "italic", size=10),
                     legend.title = element_blank()) +
  ylab("Activity") + xlab("Time (ZT)") +
  scale_color_manual(values=c("#00EE76", "#AB82FF", "darkgray"), 
                     labels = c(paste("P. greeni", "\u2642"),
                                paste("P. marginellus", "\u2642"),
                                paste("P. obscurellus", "\u2642"))) +
  scale_fill_manual(values=c("#00EE76", "#AB82FF", "gray"),
                    labels = c(paste("P. greeni", "\u2642"),
                               paste("P. marginellus", "\u2642"),
                               paste("P. obscurellus", "\u2642"))) + ylim(c(0,50))
p_photinus_activity
#ggsave("activity_ld_photinus.png",p_photinus_activity, height=4, width = 8, units=c("in"))


```

Calculate peak activity phase and nocturnality of Photinus. Plot results.
```{r}
#generate a dataframe to record information about peak activity phase and
#ratio of nocturnal/diurnal activity
diel_act_photinus = data.frame(expand.grid(uid = iLAM_photinus[meta=T]$uid,
                        day = seq(0,2,1)),
                        species = iLAM_photinus[meta=T]$species,
                        phase = 0,
                        activity_day = 0,
                        activity_night = 0,
                        before_4hr = 0,
                        before_8hr = 0,
                        before_2hr_am = 0,
                        before_4hr_am = 0) %>% arrange(uid, day)

#Set up butterworth filter, which we'll apply to each subset of binned data
#order 2, 10 Hz low-pass filter t <- seq(0, 1, len = 96) # 1 second sample of time values
bf <- butter(n = 2, W = 0.1, type = "low", plane = "z")

#return parameters for each individual cage (uid)
for (i in 1:nrow(diel_act_photinus)){
  
  sub_iLAM_photinus = iLAM_photinus[xmv(uid) == diel_act_photinus$uid[i]]
  
  #calculate phase of peak evening activity and day/night activity for each day of recording
  
  #now, isolate a specific day and calculate phase, total day activity, total night activity
  small_dt_photinus <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24)&
                                           t<=hours(diel_act_photinus$day[i]*24+24)]
  small_dt_photinus$t <- small_dt_photinus$t/60/60-diel_act_photinus$day[i]*24
  
  #for each uid, apply butterworth filter
  small_dt_photinus[, activity_butter := round(filtfilt(bf, activity_bin), 0)]
  #return phase for each uid
  small_df_photinus <- as.data.frame(rejoin(small_dt_photinus))
  
  # Returns a matrix where each row represents one peak found.
  # The first column gives the height,
  # second the position/index where the maximum is reached, 
  # third and forth the indices of where the peak begins and ends,
  # in the sense of where the pattern starts and ends.
  
  diel_act_photinus$phase[i] = small_df_photinus$t[findpeaks(small_df_photinus$activity_butter,
                                                       sortstr = TRUE,
                                                       npeaks = 4,
                                                       nups = 2)[1,2]]
  
  #calculate ratio night/day activity for each day
  day <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24) & t<hours(diel_act_photinus$day[i]*24+16)]
  
  diel_act_photinus$activity_day[i] = sum(day$activity_bin, na.rm = TRUE)
  
  night = sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24+16) & 
                              t<hours(diel_act_photinus$day[i]*24+24)]
  
  diel_act_photinus$activity_night[i] = sum(night$activity_bin, na.rm = TRUE)
  
  #calculate the evening anticipation/day (ratio activity 4 hrs from night transition/8 hrs from night transition)
   before_4hr <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24+12) & #12 for evening or morning antic
                              t<=hours(diel_act_photinus$day[i]*24+16)] #16 for evening or morning antic
   
   before_8hr <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24+8) & #8 for evening or morning antic
                              t<=hours(diel_act_photinus$day[i]*24+16)] #16 for evening or morning antic
   
  
  diel_act_photinus$before_4hr[i] = sum(before_4hr$activity_bin, na.rm = TRUE)
  diel_act_photinus$before_8hr[i] = sum(before_8hr$activity_bin, na.rm = TRUE)
  
    #calculate the morning anticipation/day (ratio activity 2 hrs from morning transition/4 hrs from morning transition)
   before_2hr_am <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24+22) & #12 for evening or morning antic
                              t<=hours(diel_act_photinus$day[i]*24+24)] #16 for evening or morning antic
   
   before_4hr_am <- sub_iLAM_photinus[t>=hours(diel_act_photinus$day[i]*24+20) & #8 for evening or morning antic
                              t<=hours(diel_act_photinus$day[i]*24+24)] #16 for evening or morning antic
   
  
  diel_act_photinus$before_2hr_am[i] = sum(before_2hr_am$activity_bin, na.rm = TRUE)
  diel_act_photinus$before_4hr_am[i] = sum(before_4hr_am$activity_bin, na.rm = TRUE)
  
  

}

diel_act_photinus$perc_noct = round(diel_act_photinus$activity_night/
                                      (diel_act_photinus$activity_night+diel_act_photinus$activity_day)*100,
                                    2)
diel_act_photinus$phase = round(as.numeric(diel_act_photinus$phase),2)
diel_act_photinus$evening_antic = round((diel_act_photinus$before_4hr/diel_act_photinus$before_8hr),2)
diel_act_photinus$morning_antic = round((diel_act_photinus$before_2hr_am/diel_act_photinus$before_4hr_am),2)


#summarize mean for each uid! #dplyr::filter(diel_act_photinus, day==2)
mean_diel_act_photinus = diel_act_photinus %>% group_by(species, day) %>%
  summarise(phase = mean(phase),
            n = n(),
            perc_noct = mean(perc_noct),
            evening_antic = mean(evening_antic, na.rm = TRUE),
            morning_antic = mean(morning_antic, na.rm = TRUE))
#summarise mean for each species
species_diel_act_photinus = mean_diel_act_photinus %>% group_by(species) %>% 
  summarise(activity_phase = round(mean(phase),2),
            ymin_phase = tryCatch(round(t.test(phase)$conf.int[1],2),
                                  error = function(e) return(NA)),
            ymax_phase = tryCatch(round(t.test(phase)$conf.int[2],2),
                                  error = function(e) return(NA)),
            n = n(),
            index_noct = mean(perc_noct),
            ymin_noct = round(t.test(perc_noct)$conf.int[1],2),
            ymax_noct = round(t.test(perc_noct)$conf.int[2],2),
            index_antic = round(mean(evening_antic),3),
            ymin_antic = round(t.test(evening_antic)$conf.int[1], 3),
            ymax_antic = round(t.test(evening_antic)$conf.int[2],3),
            index_antic_am = round(mean(morning_antic),2),
            ymin_antic_am = round(t.test(morning_antic)$conf.int[1], 2),
            ymax_antic_am = round(t.test(morning_antic)$conf.int[2], 2))

species_diel_act_photinus$ymax_noct[species_diel_act_photinus$ymax_noct>=100] = 100
species_diel_act_photinus$ymax_antic[species_diel_act_photinus$ymax_antic>=1] = 1

p_antic = ggplot(data=mean_diel_act_photinus, aes(x = species, y=evening_antic, color=species)) +
  geom_point(size=3, alpha=0.5) +
    geom_errorbar(data = species_diel_act_photinus,
                  aes(x=species, ymin=ymin_antic, ymax=ymax_antic, color=species),
                color = "black", inherit.aes = FALSE,
                size = 0.2, width=0.25) +
  scale_color_manual(values=c("#00EE76", "#AB82FF", "darkgray")) +
  theme_bw() + ylim(c(0,1)) +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(face = "italic")) +
  ylab("Anticipation Index") + xlab("Species")

p_nd = ggplot(data=mean_diel_act_photinus, aes(x = species, y=perc_noct, color=species)) +
  geom_point(size=3, alpha=0.5) +
  geom_errorbar(data = species_diel_act_photinus,
                aes(x=species, ymin=ymin_noct, ymax=ymax_noct, color=species),
                color = "black", inherit.aes = FALSE,
                size = 0.2, width=0.25) +
  scale_color_manual(values=c("#00EE76", "#AB82FF", "darkgray")) +
  theme_bw() + ylim(c(0,100)) +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(face = "italic")) +
  ylab("Nocturnality (%)") + xlab("Species")

p_all = ggarrange(p_photinus_activity,                                               
          ggarrange(p_antic, p_nd, ncol = 2, labels = c("B", "C")),
          nrow = 2,
          labels = "A") 
p_all
#ggsave("analysis_photinus.png",p_all, height=8, width = 16, units=c("in"))

```

Read in the iLAM output for Ostrinia cages. Plot activity. 
```{r}
metadata_o = read_delim("metadata_onub.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

#link entire metadata with the iLAM files
metadata_o = metadata_o %>% link_dam_metadata(result_dir = getwd()) 

iLAM_onub <- load_dam(metadata_o) 
summary(iLAM_onub)

#split data by strain: iLAM_BE vs. iLAM_UZ
iLAM_BE = iLAM_onub[xmv(strain) == "be"]
iLAM_UZ = iLAM_onub[xmv(strain) == "uz"]

#initially, work with BE data
#add uid for specific identification for each cage/experiment/species
iLAM_BE[, uid := 1:.N, meta=T]

iLAM_BE[, activity_bin := rollapply(activity, list(seq(0, 14)),
                                          FUN=function(x) sum(x, na.rm=TRUE),
                                          partial = TRUE, fill = NA),
              by = xmv(uid)]

#For each uid, retains only the 15th row (t) = 30min bins, 5 = 10min bins
iLAM_BE = iLAM_BE[, .SD[seq(1, .N, 15)], by = id]

#Calculate average activity across all bins to normalize activity_bins between cages
iLAM_BE[, uid := 1:.N, meta=T]
iLAM_BE[, fctr := .(10/mean(activity_bin)), by = xmv(uid)]
iLAM_BE[, .N, by=xmv(uid)] #verify that proper number of bins were kept

#Multiple activity_bin by normalization factor
iLAM_BE[, acn := activity_bin*fctr]


#visualize average across individuals for first 72 hours
p_be_activity = ggetho(iLAM_BE, aes(y=acn, colour=sex), time_wrap = hours(24.1)) +
  stat_pop_etho(size=0.75) +
  stat_ld_annotations(l_duration = as.numeric(hours(16)),
                      ld_colours = c("#FDFD66","black")) +
  stat_ld_annotations(x_limits = days(c(3,7)),
                      ld_colours = c("grey", "black")) +
  theme_bw() + theme(text = element_text(size = 12, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.12,0.7),
                     legend.text = element_text(size=10, face = "italic"),
                     legend.title = element_blank()) +
  ylab("Activity") + xlab("Time (ZT)") +
  scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c(paste("O. nubilalis", "\u2640"),
                                paste("O. nubilalis", "\u2642"))) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),
                    labels = c(paste("O. nubilalis", "\u2640"),
                                paste("O. nubilalis", "\u2642"))) +
  ylim(c(0,50))

p_be_activity
```

Read in the TriKinetics DAM output for Ostrinia tubes. Plot activity. 
```{r}
metadata_dam = read_delim("metadata_dam_202305.csv", 
                      delim = ",", escape_double = FALSE, trim_ws = TRUE)

metadata_dam$sex = as.factor(metadata_dam$sex)
table(metadata_dam$sex)
str(metadata_dam)

metadata_dam <- metadata_dam %>%
  link_dam_metadata(result_dir = getwd())

dam_o <- load_dam(metadata_dam) 

summary(dam_o)
dam_o[meta=T]

dam_o <- dam_o[(xmv(sex) == "male" | xmv(sex) == "female")]
summary(dam_o)

dam_o[, uid := 1:.N, meta=T]

dam_o = dam_o[t>=days(1)]
dam_o$t = dam_o$t-days(1)



dam_o[, activity_bin := rollapply(activity, list(seq(0, 14)),
                                          FUN=function(x) sum(x, na.rm=TRUE),
                                          partial = TRUE, fill = NA),
              by = xmv(uid)]

#For each uid, retains only the 15th row (t) = 30min bins, 5 = 10min bins
dam_o = dam_o[, .SD[seq(1, .N, 15)], by = id]

#Calculate average activity across all bins to normalize activity_bins between cages
dam_o[, uid := 1:.N, meta=T]
dam_o[, .N, by=xmv(uid)] #verify that proper number of bins were kept
dam_o[, .(mean_acti = 1/mean(activity_bin)), by=xmv(uid)]

dam_o[ , .(ind_sum = sum(activity_bin)), by = xmv(uid)]$ind_sum %>% median()
dam_o[, inactive := .(sum(activity_bin)<750), by = xmv(uid)] #remove wonky individuals with abnormally low mvmnt (likely dead)

dam_o <- dam_o[inactive == FALSE]
summary(dam_o)

p_dam_activity = ggetho(dam_o, aes(y=acn, colour=sex), time_wrap = hours(24.1)) +
  stat_pop_etho(size=0.70) +
  stat_ld_annotations(l_duration = as.numeric(hours(16)),
                      ld_colours = c("#FDFD66","black")) +
  stat_ld_annotations(x_limits = days(c(3,4)),
                      ld_colours = c("grey", "black")) +
  theme_bw() + theme(text = element_text(size = 12, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.12,0.70),
                     legend.text = element_text(size=10, face = "italic"),
                     legend.title = element_blank()) +
  ylab("Activity") + xlab("Time (AZT)") +
  scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c(paste("O. nubilalis", "\u2640"),
                                paste("O. nubilalis", "\u2642"))) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),
                    labels = c(paste("O. nubilalis", "\u2640"),
                               paste("O. nubilalis", "\u2642"))) + ylim(c(0,50))

p_dam_activity
```

Combine Photinus (iLAM), Ostrinia (iLAM), and Ostrinia (DAM) activity plots
```{r}
p_activity_combined = ggarrange(p_photinus_activity,
                                p_be_activity,
                                p_dam_activity,
                                nrow = 3,
                                labels = c("A", "B", "C")) 
p_activity_combined


#ggsave("activity_3x_combined.png",p_activity_combined, height=7, width = 7, units=c("in"))

```


Calculate peak activity phase and nocturnality for Ostrinia (DAM)
```{r}
#generate a datafarme to record information about peak activity phase and
#ratio of nocturnal/diurnal activity
diel_act_dam = data.frame(expand.grid(uid = dam_o[meta=T]$uid,
                        day = 0:1),
                        sex = dam_o[meta=T]$sex,
                        phase = 0,
                        activity_day = 0,
                        activity_night = 0) %>% arrange(uid, day)

#Set up butterworth filter, which we'll apply to each subset of binned data
#order 2, 10 Hz low-pass filter t <- seq(0, 1, len = 96) # 1 second sample of time values
bf <- butter(n = 2, W = 0.1, type = "low", plane = "z")

#return parameters for each individual cage (uid)
for (i in 1:nrow(diel_act_dam)){
  
  sub_dam_o = dam_o[xmv(uid) == diel_act_dam$uid[i]]
  
  #calculate phase of peak evening activity and day/night activity for each day of recording
  
  #now, isolate a specific day and calculate phase, total day activity, total night activity
  small_dt_dam <- sub_dam_o[t>=hours(diel_act_dam$day[i]*24) & t<=hours(diel_act_dam$day[i]*24+24)]
  small_dt_dam$t <- small_dt_dam$t/60/60-diel_act_dam$day[i]*24
  
  #for each uid, apply butterworth filter
  small_dt_dam[, activity_butter := round(filtfilt(bf, activity_bin), 0)]
  #return phase for each uid
  small_df_dam <- as.data.frame(rejoin(small_dt_dam))
  
  # Returns a matrix where each row represents one peak found.
  # The first column gives the height,
  # second the position/index where the maximum is reached, 
  # third and forth the indices of where the peak begins and ends,
  # in the sense of where the pattern starts and ends.
  
  peak = pracma::findpeaks(small_df_dam$activity_butter,
                   sortstr = TRUE,
                   npeaks = 3,
                   nups = 2)[1,2]
  
  if(is.null(peak)){
    diel_act_dam$phase[i] <- small_df_dam$t[which.max(small_df_dam$activity_butter)]
  } else{
    diel_act_dam$phase[i] = small_df_dam$t[peak]
  }
  
  
  #calculate ratio night/day activity for each day
  day <- sub_dam_o[t>=hours(diel_act_dam$day[i]*24) & t<=hours(diel_act_dam$day[i]*24+16)]
  diel_act_dam$activity_day[i] = sum(day$activity, na.rm = TRUE) #to control for photophase length
  night = sub_dam_o[t>=hours(diel_act_dam$day[i]*24+16) & t<=hours(diel_act_dam$day[i]*24+24)]
  diel_act_dam$activity_night[i] = sum(night$activity, na.rm = TRUE) #to control for scotophase length
  
}

diel_act_dam$perc_noct = diel_act_dam$activity_night/(diel_act_dam$activity_night+diel_act_dam$activity_day)*100
diel_act_dam$phase = round(as.numeric(diel_act_dam$phase),2)
diel_act_dam$perc_noct = round(as.numeric(diel_act_dam$perc_noct),2)

diel_act_dam = diel_act_dam %>% group_by(uid, sex) %>%
  summarise(phase = mean(phase),
            activity_day = mean(activity_day),
            activity_night = mean(activity_night),
            perc_noct = mean(perc_noct)) %>% drop_na()

#summarise mean for each sex
sex_diel_act_dam = diel_act_dam %>% group_by(sex) %>% 
  summarise(activity_phase = mean(phase, na.rm=TRUE),
            n = n(),
            ymin_phase = t.test(phase)$conf.int[1],
            ymax_phase =  t.test(phase)$conf.int[2],
            index_noct = mean(perc_noct, na.rm = TRUE),
            ymin_noct = t.test(perc_noct)$conf.int[1],
            ymax_noct = t.test(perc_noct)$conf.int[2])

sex_diel_act_dam$ymax_noct[sex_diel_act_dam$ymax_noct>100] = 100
sex_diel_act_dam$ymax_phase[sex_diel_act_dam$ymax_phase>24] = 24

p_dam_phase = ggplot(data=diel_act_dam, aes(x=sex, y=phase, color=sex)) +
  geom_point(size=3, alpha=0.5) +
  geom_errorbar(data = sex_diel_act_dam,
                aes(x=sex, ymin=ymin_phase,ymax=ymax_phase, color=sex),
                 color = "black", inherit.aes = FALSE,
                 size = 0.2, width=0.25) +
    scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c("Female", "Male")) +
  ylim(c(13,23)) + coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.y = element_blank()) +
  ylab("Peak Activity (ZT)")

p_dam_nd = ggplot(data=diel_act_dam, aes(x = sex, y=perc_noct, color=sex)) +
  geom_point(size=3, alpha=0.5) +
  geom_errorbar(data = sex_diel_act_dam,
                aes(x=sex, ymin=ymin_noct,
                    ymax=ymax_noct, color=sex),
                color = "black", inherit.aes = FALSE,
                size = 0.2, width=0.25) +
 scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c("Female", "Male")) +
  theme_bw() + ylim(c(0,100)) +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank()) +
  ylab("Nocturnality (%)")

p_dam_all = ggarrange(p_dam_activity,                                               
          ggarrange(p_dam_phase, p_dam_nd, ncol = 2, labels = c("B", "C")),
          nrow = 2,
          labels = "A") 
p_dam_all
#ggsave("analysis_dam_activity.png",p_dam_all, height=8, width = 16, units=c("in"))


```

Calculate peak activity phase and nocturnality for Ostrinia (iLAM)
```{r}
#generate a datafarme to record information about peak activity phase and
#ratio of nocturnal/diurnal activity
diel_act_be = data.frame(expand.grid(uid = iLAM_BE[meta=T]$uid,
                        day = 0:2),
                        sex = iLAM_BE[meta=T]$sex,
                        phase = 0,
                        activity_day = 0,
                        activity_night = 0) %>% arrange(uid, day)

#Set up butterworth filter, which we'll apply to each subset of binned data
#order 2, 10 Hz low-pass filter t <- seq(0, 1, len = 96) # 1 second sample of time values
bf <- butter(n = 2, W = 0.1, type = "low", plane = "z")

#return parameters for each individual cage (uid)
for (i in 1:nrow(diel_act_be)){
  
  sub_iLAM_be = iLAM_BE[xmv(uid) == diel_act_be$uid[i]]
  
  #calculate phase of peak evening activity and day/night activity for each day of recording
  
  #now, isolate a specific day and calculate phase, total day activity, total night activity
  small_dt_be <- sub_iLAM_be[t>=hours(diel_act_be$day[i]*24) & t<=hours(diel_act_be$day[i]*24+24)]
  small_dt_be$t <- small_dt_be$t/60/60-diel_act_be$day[i]*24
  
  #for each uid, apply butterworth filter
  small_dt_be[, activity_butter := round(filtfilt(bf, activity_bin), 0)]
  #return phase for each uid
  small_df_be <- as.data.frame(rejoin(small_dt_be))
  
  # Returns a matrix where each row represents one peak found.
  # The first column gives the height,
  # second the position/index where the maximum is reached, 
  # third and forth the indices of where the peak begins and ends,
  # in the sense of where the pattern starts and ends.
  
   peak = findpeaks(small_df_be$activity_butter,
                   sortstr = TRUE,
                   npeaks = 5,
                   nups = 2)[1,2]
  
  if(is.null(peak)){
    diel_act_be$phase[i] <- NA
  } else{
    diel_act_be$phase[i] = small_df_be$t[peak]
  }
  
  #calculate ratio night/day activity for each day
  day <- sub_iLAM_be[t>=hours(diel_act_be$day[i]*24) & t<=hours(diel_act_be$day[i]*24+16)]
  diel_act_be$activity_day[i] = sum(day$activity_bin, na.rm = TRUE) #to control for photophase length
  night = sub_iLAM_be[t>=hours(diel_act_be$day[i]*24+16) & t<=hours(diel_act_be$day[i]*24+24)]
  diel_act_be$activity_night[i] = sum(night$activity_bin, na.rm = TRUE) #to control for scotophase length
  
}

diel_act_be$perc_noct = diel_act_be$activity_night/(diel_act_be$activity_night+diel_act_be$activity_day)*100
diel_act_be$phase = round(as.numeric(diel_act_be$phase),2)
diel_act_be$perc_noct = round(as.numeric(diel_act_be$perc_noct),2)

#summarize mean for each uid!
mean_diel_act_be = diel_act_be %>% group_by(sex, day) %>%
  summarise(phase = mean(phase),
            n = n(),
            perc_noct = mean(perc_noct))
#summarise mean for each sex
sex_diel_act_be = mean_diel_act_be %>% group_by(sex) %>% 
  summarise(activity_phase = mean(phase),
            n = n(),
            ymin_phase = t.test(phase)$conf.int[1],
            ymax_phase =  t.test(phase)$conf.int[2],
            index_noct = mean(perc_noct),
            ymin_noct = t.test(perc_noct)$conf.int[1],
            ymax_noct = t.test(perc_noct)$conf.int[2])

sex_diel_act_be$ymax_noct[sex_diel_act_be$ymax_noct>100] = 100

p_be_phase = ggplot(data=mean_diel_act_be, aes(x=sex, y=phase, color=sex)) +
  geom_point(size=3, alpha=0.5) +
  geom_errorbar(data = sex_diel_act_be,
                aes(x=sex, ymin=ymin_phase,ymax=ymax_phase, color=sex),
                 color = "black", inherit.aes = FALSE,
                 size = 0.2, width=0.25) +
    scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c("Female", "Male")) +
  ylim(c(18,22.5)) + coord_flip() +
  theme_bw() +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.y = element_blank()) +
  ylab("Peak Activity (ZT)")

p_be_nd = ggplot(data=mean_diel_act_be, aes(x = sex, y=perc_noct, color=sex)) +
  geom_point(size=3, alpha=0.5) +
  geom_errorbar(data = sex_diel_act_be,
                aes(x=sex, ymin=ymin_noct,
                    ymax=ymax_noct, color=sex),
                color = "black", inherit.aes = FALSE,
                size = 0.2, width=0.25) +
 scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     labels = c("Female", "Male")) +
  theme_bw() + ylim(c(0,100)) +
  theme(text = element_text(size = 12, family = "sans", color = 'black'),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank()) +
  ylab("Nocturnality (%)")

p_be_all = ggarrange(p_be_activity,                                               
          ggarrange(p_be_phase, p_be_nd, ncol = 2, labels = c("B", "C")),
          nrow = 2,
          labels = "A") 
p_be_all
ggsave("analysis_be_activity.png",p_be_all, height=8, width = 16, units=c("in"))


mean_diel_act_be$species = "O. nubilalis"
mean_diel_act_photinus$sex = "male"

combined_results = rbind(mean_diel_act_be, mean_diel_act_photinus)
combined_results$phase = round(combined_results$phase,2)
write.csv(combined_results, "analysis_activity.csv")

combined_results_mean = rbind((sex_diel_act_be %>% mutate("species" = "O. nubilalis")),
                              (sex_diel_act_dam %>% mutate("species" = "O. nubilalis")),
                              (species_diel_act_photinus[,1:8] %>% mutate("sex" = "male")))
write.csv(combined_results_mean, "analysis_activity_mean.csv")



```

Calculate % rhythmic, rhythmic strength, and period length (tau) for bivoltine Ostrinia cages
```{r}
#Subset iLAM_BE to only retain counts recorded from days in DD
small_dt_be <- iLAM_BE[t > hours(72)]

#Calculate rhythmicity by chi_sq_periodogram method
per_iLAM_BE <- periodogram(activity_bin,
                           small_dt_be, FUN = chi_sq_periodogram,
                           resample_rate = 1/mins(10),
                           period_range=c(hours(18),hours(27)))

per_iLAM_BE <-find_peaks(per_iLAM_BE)

ggperio(per_iLAM_BE, aes(period, power, colour=sex)) + 
  stat_pop_etho() +
  geom_line(aes(y = signif_threshold), color="black", linetype = "longdash") + 
  theme_bw() + theme(text = element_text(size = 18, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.2,0.8)) +
  ylab("Power") + xlab("Period Length (hr)")


#Rhythmicity for populations are determined by Power-Significance
#If (P-S)<0, then rhythmicity<-0

# Create a data table with all_ids and missing values for the columns
summary_dt_BE <- per_iLAM_BE[meta=T]
# Get the summary data for ids that have a peak==1
summary_dt_BE_peak <- per_iLAM_BE[peak == 1,
                                  .(first_peak_period = period,
                                    first_peak_rel_power = power - signif_threshold,
                                    is_rhythmic = TRUE), 
                                  by = id]
# Left join summary_dt_BE_peak with summary_dt_BE
summary_dt_BE <- summary_dt_BE_peak[summary_dt_BE, on = "id"]

# Replace NA values in is_rhythmic with FALSE
summary_dt_BE[is.na(is_rhythmic), is_rhythmic := FALSE]
summary_dt_BE[, period := round(first_peak_period/60/60,2) ]

output_be = as.data.frame(summary_dt_BE)
ggplot(output_be, aes(sex, period, fill= sex)) + 
  geom_point(aes(size=first_peak_rel_power), alpha=1) +
  ylim(c(19,23.5)) + theme_bw()


```

Calculate % rhythmic, rhythmic strength, and period length (tau) for univoltine Ostrinia cages
```{r}
iLAM_UZ = iLAM_onub[xmv(strain) == "uz"]

#add uid for specific identification for each cage/experiment/species
iLAM_UZ[, uid := 1:.N, meta=T]

library(zoo)
iLAM_UZ[, activity_bin := rollapply(activity, list(seq(0, 14)),
                                          FUN=function(x) sum(x, na.rm=TRUE),
                                          partial = TRUE, fill = NA),
              by = xmv(uid)]

#For each uid, retains only the 15th row (t) = 30min bins, 5 = 10min bins
iLAM_UZ = iLAM_UZ[, .SD[seq(1, .N, 15)], by = id]

#Calculate average activity across all bins to normalize activity_bins between cages
iLAM_UZ[, uid := 1:.N, meta=T]
iLAM_UZ[, fctr := .(10/mean(activity_bin)), by = xmv(uid)]
iLAM_UZ[, .N, by=xmv(uid)] #verify that proper number of bins were kept

small_dt <- iLAM_UZ[t <=hours(120)] #retain first 5 days in dd

#Average double-plotted actograms for each group across the experiment
ggetho(iLAM_UZ, aes(x=t, z=activity_bin), multiplot = 2) + 
  stat_bar_tile_etho() +
  facet_wrap( ~ sex) + 
  theme_bw() + theme(text = element_text(size = 18, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.8,0.8))

#Calculate rhythmicity by chi_sq_periodogram method
per_iLAM_UZ <- periodogram(activity_bin,
                           iLAM_UZ, FUN = chi_sq_periodogram,
                           resample_rate = 1/mins(10),
                           period_range=c(hours(18),hours(27)))

per_iLAM_UZ <-find_peaks(per_iLAM_UZ)

ggperio(per_iLAM_UZ, aes(period, power, colour=sex)) + 
  stat_pop_etho() +
  geom_line(aes(y = signif_threshold), color="black", linetype = "longdash") + 
  theme_bw() + theme(text = element_text(size = 18, family = "sans", color = 'black'),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     legend.position = c(0.2,0.8)) +
  ylab("Power") + xlab("Period Length (hr)")


# Create a data table with all_ids and missing values for the columns
summary_dt_UZ <- per_iLAM_UZ[meta=T]
# Get the summary data for ids that have a peak==1
summary_dt_UZ_peak <- per_iLAM_UZ[peak == 1,
                                  .(first_peak_period = period,
                                    first_peak_rel_power = power - signif_threshold,
                                    is_rhythmic = TRUE), 
                                  by = id]
# Left join summary_dt_UZ_peak with summary_dt_UZ
summary_dt_UZ <- summary_dt_UZ_peak[summary_dt_UZ, on = "id"]

# Replace NA values in is_rhythmic with FALSE
summary_dt_UZ[is.na(is_rhythmic), is_rhythmic := FALSE]
summary_dt_UZ[, period := round(first_peak_period/60/60,2) ]

output_uz = as.data.frame(summary_dt_UZ)
ggplot(output_uz, aes(sex, period, fill= sex)) + 
  geom_point(aes(size=first_peak_rel_power), alpha=1) +
  ylim(c(19,23.5)) + theme_bw()

output_onub = rbind(output_be,output_uz) %>%
  select(-c(experiment_id, start_datetime, stop_datetime, swap_datetime, file_info, id, uid))
output_onub = data.frame(output_onub)
output_onub$sex = as.factor(output_onub$sex)
output_onub$strain = as.factor(output_onub$strain)

output_onub %>% ungroup() %>%
  group_by(sex, strain) %>%
  summarize(n_cage = n(),
            percent_rhythmic = round(length(period[!is.na(period)])/n_cage*100, 0),
            tau = mean(period, na.rm = TRUE),
            ymin_tau = ifelse(sum(!is.na(period)) >= 3, t.test(period)$conf.int[1], NA),
            ymax_tau = ifelse(sum(!is.na(period)) >= 3, t.test(period)$conf.int[2], NA),
            rhythmic_strength = round(mean(first_peak_rel_power, na.rm=TRUE),0),
            ymin_strength = ifelse(sum(!is.na(first_peak_rel_power)) >= 3,
                                   t.test(first_peak_rel_power)$conf.int[1], NA),
            ymax_strength = ifelse(sum(!is.na(first_peak_rel_power)) >= 3,
                                   t.test(first_peak_rel_power)$conf.int[2], NA))

output_onub$first_peak_rel_power = round(output_onub$first_peak_rel_power,0)

write.csv(output_onub, "analysis_circ.csv")

```


